---
title: "Introduction to counterfactuals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to counterfactuals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this document, we explain the `counterfactuals` workflow for both a classification and a regression task.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 7,
  fig.height = 3,
  comment = "#>"
)
options(width = 130)
```

```{r setup, message=FALSE}
library(counterfactuals)
library(iml)
library(randomForest)
```


## Classification Models

To demonstrate the workflow for classification tasks, we search for counterfactual explanations for diabetes tested
patients with `MOC`. 

### Data: Pima Indians Diabetes Databas

As training data we use the Pima Indians Diabetes Database from the `mlbench` package.
The data set contains 768 observations with 8 features and the binary target variable `diabetes`:

```{r, echo=FALSE}
column_descr = data.frame(
  rbind(
    cbind("pregnant", "Number of times pregnant"),
    cbind("glucose", "Plasma glucose concentration (glucose tolerance test)"),
    cbind("pressure", "Diastolic blood pressure (mm Hg)"),
    cbind("mass", "Body mass index (weight in kg/(height in m)\\^2)"),
    cbind("pedigree", "Diabetes pedigree function"),
    cbind("age", "	Age (years)"),
    cbind("diabetes", "Class variable (test for diabetes)")
  )
)
names(column_descr) <- c("Variable", "Description")
knitr::kable(column_descr, escape = FALSE, format = "html", table.attr = "style='width:100%;'")
```

```{r}
data(PimaIndiansDiabetes, package = "mlbench")  
```


Converting `integerish` features to the `integer` data type ensures that counterfactuals will
only contain `integer` values for these features (e.g. no `2.76` pregnancies).

```{r}
PimaIndiansDiabetes$pregnant = as.integer(PimaIndiansDiabetes$pregnant)
PimaIndiansDiabetes$glucose = as.integer(PimaIndiansDiabetes$glucose)
PimaIndiansDiabetes$age = as.integer(PimaIndiansDiabetes$age)
```


### Fitting a model

First we train a model to predict `diabetes`. Note that we leave out one observation from the
training data which is our `x_interest`.

```{r}
set.seed(20210716)
rf = randomForest(diabetes ~ ., data = PimaIndiansDiabetes[-154L, ])
```

### Setting up an iml::Predictor() object

An [`iml::Predictor`](https://christophm.github.io/iml/reference/Predictor.html) object serves as a wrapper for different 
model types. It contains the model and the data for its analysis.

```{r}
predictor = Predictor$new(rf, type = "prob")
```


### Find counterfactuals

For `x_interest` the model predicts:

```{r}
x_interest = PimaIndiansDiabetes[154L, ]
predictor$predict(x_interest)
```

Now we can examine which risk factors need to be changed to obtain a predicted probability of
at least 50% for a diabetes disease.

Since we want to use `MOC` for a classification model, we initialize a `MOCClassif` object.

```{r, eval=FALSE}
moc_classif = MOCClassif$new(predictor)
```

Then we use the `$find_counterfactuals()` method to find counterfactual explanations for `x_interest`. 

```{r, eval=FALSE}
cfactuals = moc_classif$find_counterfactuals(x_interest, desired_class = "pos", desired_prob = c(0.5, 1))
```

```{r, echo=FALSE, results='hide', message=FALSE}
if (!file.exists("introduction-res/cfactuals.RDS")) {
  moc_classif = MOCClassif$new(predictor)
  cfactuals = moc_classif$find_counterfactuals(x_interest, desired_class = "pos", desired_prob = c(0.5, 1))
  saveRDS(moc_classif, "introduction-res/moc_classif.RDS")
  saveRDS(cfactuals, "introduction-res/cfactuals.RDS")
}
moc_classif = readRDS("introduction-res/moc_classif.RDS")
cfactuals = readRDS("introduction-res/cfactuals.RDS")
```


### The counterfactuals object

The resulting `Counterfactuals` object holds the found counterfactuals and has several methods for their evaluation and 
visualization.

```{r}
class(cfactuals)
```


```{r}
print(cfactuals)
```

We can use the `$predict()` method to predict the outcome of the counterfactuals found.

```{r}
head(cbind(cfactuals$data, cfactuals$predict()), 5L)
```

The `$evaluate()` method returns the counterfactuals along with evaluation measures (`dist_x_interest`, `dist_target`,
`nr_changed` and `dist_train`). Setting the `show_diff` argument to `TRUE` will show the counterfactuals as their difference
to `x_interest`. `NA` means "no difference".

```{r}
head(cfactuals$evaluate(show_diff = TRUE), 5L)
```

To plot the frequency of feature changes across all counterfactuals, we can use the `$plot_freq_of_feature_changes()` method.

```{r}
cfactuals$plot_freq_of_feature_changes()
```

The parallel plot connects the (scaled) feature values of each counterfactual and highlights `x_interest` in blue.

```{r, message=FALSE}
cfactuals$plot_parallel()
```

The white dot in the prediction surface plot represents `x_interest`. All counterfactuals that differ from `x_interest` only in the 
selected features are represented as black dots.

```{r, fig.height=3.5}
cfactuals$plot_surface(feature_names = c("mass", "glucose"))
```


### MOC diagnostics

A `MOCClassif` object has several diagnostics methods.

The method `$plot_search()` plots two objective values of all individuals across generations.

```{r}
moc_classif$plot_search()
```

To plot the evolution of the mean and minimum objective values along with the dominated hypervolume, we can use the 
`$plot_statistics()` method.

```{r, fig.height=2, results='asis'}
moc_classif$plot_statistics()
```


## Regression Models

Finding counterfactuals for regression models is analogous to classification models. In this example, we search for
counterfactuals explanations for housing prices with `WhatIf`.

### Data: Boston Housing Data

As training data we use the Boston Housing dataset from the `mlbench` package. 
The data set contains 506 observations with 13 features and the continuous target variable `medv`:

```{r, echo=FALSE}
# data(BostonHousing, package = "mlbench")  
# df = data.frame(
#   Variable = colnames(BostonHousing),
#   Description = c(
#     "Per capita crime rate by town",
#     "Proportion of residential land zoned for lots over 25,000 sq.ft.",
#     "Proportion of non-retail business acres per town",
#     "Charles River dummy variable (= 1 if tract bounds river; 0 otherwise)",
#     "Nitrogen oxides concentration (parts per 10 million).",
#     "Average number of rooms per dwelling.",
#     "Proportion of owner-occupied units built prior to 1940",
#     "Weighted mean of distances to five Boston employment centres.",
#     "Index of accessibility to radial highways.",
#     "Full-value property-tax rate per $10,000.",
#     "Pupil-teacher ratio by town.",
#     "1000(Bk - 0.63)^2 where Bk is the proportion of blacks by town.",
#     "Percentage of lower status of the population",
#     "Median value of owner-occupied homes in USD 1000's"
#   )
# )
# knitr::kable(df, format = "html", table.attr = "style='width:100%;'")
```

```{r}
data(BostonHousing, package = "mlbench")  
```


### Fitting a model

First we train a model to predict `medv`. Note that we leave out one observation from the
training data which is our `x_interest`.

```{r}
set.seed(20210713)
rf = randomForest(medv ~ ., data = BostonHousing[-1L, ])
```

### Setting up an iml::Predictor() object

Then we initialize an [`iml::Predictor`](https://christophm.github.io/iml/reference/Predictor.html) object.

```{r}
predictor = Predictor$new(rf)
```


### Find counterfactuals

As `x_interest` we again simply use an observation from our training data.

```{r}
x_interest = BostonHousing[1L, ]
predictor$predict(x_interest)
```

Since we want to use `WhatIf` for a regression model, we initialize a `WhatIfRegr` object.

```{r}
whatif_regr = WhatIfRegr$new(predictor, n_counterfactuals = 5L)
```

Then we use the `$find_counterfactuals()` method to find counterfactual explanations for `x_interest`. Note that for
regression tasks we only have to specifiy `desired_outcome` compared to `desired_class` and `desired_probs` for
classification tasks.
 
```{r}
cfactuals = whatif_regr$find_counterfactuals(x_interest, desired_outcome = c(30, 32))
```

As a result, we obtain a `Counterfactuals` object just like for the classification task.

```{r}
cfactuals
```

