---
title: "counterfactuals R package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{counterfactuals R package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In the following, we explain the `counterfactuals` workflow for both a classification and a regression task using
concrete use cases.

```{r, include = FALSE}
# NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 7,
  fig.height = 3,
  comment = "#>"
  # purl = NOT_CRAN,
  # eval = NOT_CRAN
)
options(width = 200)
```

```{r setup, message=FALSE}
library("counterfactuals")
library("iml")
library("randomForest")
library("mlr3")
library("mlr3learners")
library("data.table")
```


## Classification Task

To illustrate the `counterfactuals` workflow for classification tasks, we search 
for counterfactuals with `MOC` (Dandl et al. 2020) for a bank customer, whose 
credit application is rejected.

### Data: German Credit Dataset

As training data, we use the German Credit Data from the `rchallenge`` package.
The data set contains 1000 observations with 21 features and the binary target variable `credit_risk`.
For illustrative purposes, we only consider 8 of the 21 features in the following: 

```{r}
data(german, package = "rchallenge")  
credit = german[, c("duration", "amount", "purpose", "age", "savings", "employment_duration", "housing", "number_credits", "credit_risk")]
```

```{r, echo=FALSE}
column_descr = data.frame(
  rbind(
    cbind("duration", "Credit duration"),
    cbind("amount", "Credit amount (DM)"),
    cbind("purpose", "Purpose of credit"),
    cbind("age", "Age (years)"),
    cbind("savings", "Amount in savings account"),
    cbind("employment_duration", "Years in present employment"),
    cbind("housing", "Type of housing"),
    cbind("number_credits", "Number of credits"),
    cbind("credit_risk", "Class variable (credit worthiness)")
  )
)
names(column_descr) <- c("Variable", "Description")
knitr::kable(column_descr, escape = FALSE, format = "html", table.attr = "style='width:100%;'")
```

### Fitting a model

First, we train a model to predict whether a credit is `good` or `bad`, omitting one observation from the training data, which is `x_interest`.

```{r}
set.seed(20210816)
rf = randomForest::randomForest(credit_risk ~ ., data = credit[-998L,])
```

### Setting up an iml::Predictor() object

An [`iml::Predictor`](https://christophm.github.io/iml/reference/Predictor.html) object serves as a wrapper for different 
model types. It contains the model and the data for its analysis.

```{r}
predictor = iml::Predictor$new(rf, type = "prob")
x_interest = credit[998L, ]
predictor$predict(x_interest)
```

For `x_interest`, the model predicts a probability of being a bad credit risk of `r round(predictor$predict(x_interest), 2)[2]`.

### Find counterfactuals

Now, we examine which factors need to be changed to incrase the predicted probability of being a good credit risk to more than 60%.\
Since we want to apply MOC to a classification model, we initialize a `MOCClassif` object.
Individuals whose prediction is farther away from the desired prediction than `epsilon` can be penalized.
Here, we set `epsilon = 0`, penalizing all individuals whose prediction is outside the desired interval. 
With the `fixed_features` argument, we can fix non-actionable features, here `age` and `employment_duration`.

```{r, eval=FALSE}
moc_classif = MOCClassif$new(
  predictor, epsilon = 0, fixed_features = c("age", "employment_duration"), init_strategy = "icecurve", quiet = TRUE, termination_crit = "genstag", n_generations = 10L
)
```

Then, we use the `find_counterfactuals()` method to find counterfactuals for `x_interest`. 
As we aim to find counterfactuals with a predicted probability of being a good credit risk of at least 60%, we set the `desired_class` to 
`"good"` and the `desired_prob` to `c(0.6, 1)`.

```{r, eval=FALSE}
cfactuals = moc_classif$find_counterfactuals(
  x_interest, desired_class = "good", desired_prob = c(0.6, 1)
)
```

```{r, echo=FALSE, results='hide', message=FALSE}
if (!file.exists("introduction-res/cfactuals_credit.RDS")) {
  moc_classif = MOCClassif$new(
  predictor, epsilon = 0, fixed_features = c("age", "employment_duration"), init_strategy = "icecurve", termination_crit = "genstag", n_generations = 10L, quiet = TRUE
)
  cfactuals = moc_classif$find_counterfactuals(x_interest, desired_class = "good", desired_prob = c(0.6, 1))
  dir.create("introduction-res")
  saveRDS(moc_classif, file = "introduction-res/moc_classif_credit.RDS")
  saveRDS(cfactuals, file = "introduction-res/cfactuals_credit.RDS")
}
moc_classif = readRDS("introduction-res/moc_classif_credit.RDS")
cfactuals = readRDS("introduction-res/cfactuals_credit.RDS")
```


### The counterfactuals object

The resulting `Counterfactuals` object holds the counterfactuals in the `data` field and possesses several methods for their 
evaluation and visualization.

```{r}
class(cfactuals)
```

Printing a `Counterfactuals` object, gives an overview of the results.
```{r}
print(cfactuals)
```

The `predict()` method returns the predictions for the counterfactuals.

```{r}
head(cbind(cfactuals$data, cfactuals$predict()), 5L)
```

The `evaluate()` method returns the counterfactuals along with the evaluation measures `dist_x_interest`, `dist_target`,
`no_changed`, and `dist_train`. \
Setting the `show_diff` argument to `TRUE` displays the counterfactuals as their difference
to `x_interest`: for a numeric feature, positive values indicate an increase compared to the feature value in `x_interest`
and negative values indicate a decrease; for factors, the counterfactual feature value is displayed if it 
differs from `x_interest.`; `NA` means "no difference" in both cases.

```{r}
head(cfactuals$evaluate(show_diff = TRUE, measures = c("dist_x_interest", "dist_target", "no_changed", "dist_train")), 5L)
```

By design, not all counterfactuals generated with MOC have a prediction equal to the desired 
prediction. We can use `subset_to_valid()` to omit all counterfactuals that do not achieve 
the desired predicition. 
```{r}
cfactuals$subset_to_valid()
cfactuals
```

The `plot_freq_of_feature_changes()` method plots the frequency of feature changes across all
remaining counterfactuals. 
Setting `subset_zero = TRUE` removes all unchanged features from the plot.

```{r, fig.height=2}
cfactuals$plot_freq_of_feature_changes(subset_zero = TRUE)
```

The parallel plot connects the (scaled) feature values of each counterfactual and highlights `x_interest` in blue.

```{r, message=FALSE, fig.height=2.5}
cfactuals$plot_parallel(feature_names = c("duration", "amount"), row_ids = NULL, 
 digits_min_max = 2L)
```

The white dot in the prediction surface plot represents `x_interest`. All counterfactuals that differ from `x_interest` only in the 
selected features are displayed as black dots. The tick marks next to the axes indicate the marginal distribution of the
counterfactuals.


```{r, fig.height=3}
cfactuals$plot_surface(feature_names = c("duration", "amount"))
```


### MOC diagnostics

Additional diagnostic tools for MOC are available as part of the MOCClassif
and MOCRegr class. 
For example, the hypervolume indicator (Zitzler and Thiele 1998) given a 
reference point (that represents the maximal values of the objectives) could be computed.
The evolution of the hypervolume indicator can be plotted together with the evolution of mean and minimum objective values 
using the `plot_statistics()` method.

```{r, fig.height=2.5, results='asis', eval=TRUE}
moc_classif$plot_statistics(centered_obj = TRUE)
```

Ideally, one would like the mean value of each objective to decrease over the generations, leading to an increase of the
hypervolume.
We could visualize the objective values of the emerging candidates throughout the generations via 
the  `plot_search` method for pairs of objectives. 
```{r, fig.height=2.5}
moc_classif$plot_search(objectives = c("dist_train", "dist_target"))
moc_classif$plot_search(objectives = c("dist_x_interest", "dist_train"))
```


## Regression Tasks

Finding counterfactuals for regression models is analogous to classification models. In this example, we use 
`NICE` (Brughmans et al. (2022)) to search for counterfactuals for plasma . 
Brughmans et al. introduced `NICE` only for the classification setting but for 
this package the method was extended to also work for regression tasks by allowing 
prediction functions to return real-valued outcomes instead of classification scores. 

### Data: Plasma Retinol

As training data, we use the plasma dataset from the `gamlss.data` package. 
The dataset contains 315 observations with 13 features and the (continuous) target variable `retplasma`, the plasma retinol concentration in ng/ml.
The plasma retinol concentration is interesting because low concentrations are associated with an increased risk for some types of cancer (Harrell 2022). 

```{r}
data(plasma, package = "gamlss.data")
```

```{r, echo=FALSE}
df = data.frame(
  Variable = colnames(plasma),
  Description = c(
    "Age (years)",
    "Sex (1 = male, 2 = female)",
    "Smoking status (1 = never, 2 = former, 3 = current smoker)",
    "Body mass index (weight/(height^2))",
    "Vitamin use (1 = yes, fairly often, 2 = yes, not often, 3 = no)",
    "Number of calories consumed per day",
    "Grams of fat consumed per day",
    "Grams of fiber consumed per day",
    "Number of alcoholic drinks consumed per week",
    "Cholesterol consumed (mg per day)",
    "Dietary beta-carotene consumed (mcg per day)",
    "Dietary retinol consumed (mcg per day)",
    "Plasma beta-carotene (ng/ml)",
    "Plasma retinol (ng/ml)"
  )
)
knitr::kable(df, format = "html", table.attr = "style='width:100%;'")
```

### Fitting a model

First, we train a model to predict `plasma_retinol`, again omitting `x_interest` from the training data.
This time we use a regression tree trained with the `mlr3` and `rpart` package. 

```{r}
tsk = mlr3::TaskRegr$new(id = "plasma", backend = plasma[-100L,], target = "retplasma")
tree = lrn("regr.rpart")
model = tree$train(tsk)
```

### Setting up an iml::Predictor() object

Then, we initialize an [`iml::Predictor`](https://christophm.github.io/iml/reference/Predictor.html) object.

```{r}
predictor = iml::Predictor$new(model, data = plasma, y = "retplasma")
x_interest = plasma[100L, ]
predictor$predict(x_interest)
```

For `x_interest`, the model predicts a plasma concentration of `r round(predictor$predict(x_interest), 2)[1]`.

### Find counterfactuals

Since we want to apply `NICE` to a regression model, we initialize a `NICERegr` object. 
For regression models, we define a correctly predicted datapoint when its prediction
is less than a user-specified value away. Here we allow for a deviation of 
`margin_correct = 0.5`. In this example, we aim for proximal counterfactuals 
in additional to sparse ones, such that we set `optimization = "proximity"`. 

```{r}
nice_regr = NICERegr$new(predictor, optimization = "proximity", 
  margin_correct = 0.5, return_multiple = FALSE)
```

Then, we use the `find_counterfactuals()` method to find counterfactuals for `x_interest` with a predicted
plasma concentration in the interval [500, Inf].
 
```{r}
cfactuals = nice_regr$find_counterfactuals(x_interest, desired_outcome = c(500, Inf))
```

### The counterfactuals object

As a result, we obtain a `Counterfactuals` object, just like for the classification task.

```{r}
cfactuals
```

### User-defined distance function
At the beginning, `NICE` calculates the distance of `x_interest` to each of the 
training samples. By default, Gower's distance measures this but users could 
also specify their own distance functions in the `distance_function` argument. 
For example, the Gower distance can be replaces by the L_0 norm. 

```{r}
l0_norm = function(x, y, data) {
  res = matrix(NA, nrow = nrow(x), ncol = nrow(y))
  for (i in seq_len(nrow(x))) {
    for (j in seq_len(nrow(y))) {
      res[i, j] = sum(x[i,] != y[j,])
    }
  }
  res
}
```

A short example illustrates the functionality of `l0_norm()`. 

```{r}
xt = data.table::data.table(a = c(0.5), b = c("a"))
yt = data.table::data.table(a = c(0.5, 3.2, 0.1), b = c("a", "b", "a"))
l0_norm(xt, yt, data = NULL)
```

Replacing the distance function is fairly easy: 

```{r}
nice_regr = NICERegr$new(predictor, optimization = "proximity", 
  margin_correct = 0.5, return_multiple = FALSE, 
  distance_function = l0_norm)
cfactuals = nice_regr$find_counterfactuals(x_interest, desired_outcome = c(500, 1000))
cfactuals
```

# References

Dandl, Susanne, Christoph Molnar, Martin Binder, and Bernd Bischl. 2020. “Multi-Objective Counterfactual Explanations.” In Parallel Problem Solving from Nature – PPSN XVI, edited by Thomas Bäck, Mike Preuss, André Deutz, Hao Wang, Carola Doerr, Michael Emmerich, and Heike Trautmann, 448–469. Cham: Springer International Publishing. \doi{10.1007/978-3-030-58112-1_31}.

Brughmans D, Martens D (2022). “NICE: An Algorithm for Nearest Instance Counterfactual
Explanations.” Technical report, <arXiv:2104.07411> v2.

Zitzler, Eckart, and Lothar Thiele. 1998. “Multiobjective Optimization Using Evolutionary Algorithms—a Comparative Case Study.” In International Conference on Parallel Problem Solving from Nature, 292–301. Springer.

Harrell, F. E. 2002. “Plasma Retinol and Beta-Carotene Dataset“. https://hbiostat.org/data/repo/plasma.html.
