---
title: "Other types of models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Other types of models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In the following, we explain the `counterfactuals` workflow for both a classification and a regression task using
concrete use cases.

```{r, echo = FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 7,
  fig.height = 3,
  comment = "#>",
  purl = NOT_CRAN,
  eval = NOT_CRAN
)
options(width = 200)
```

```{r pkg, message=FALSE}
library("counterfactuals")
library("iml")
library("rpart")
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Other types of models

The `Predictor` class of the `iml` package provides the necessary flexibility 
to cover classification and regression models fitted with diverse R packages. 
In the introduction vignette, we saw models fitted with the `mlr3` and `randomForest` packages.
In the following, we show extensions to - an classification tree fitted with 
the `caret` package, the `mlr` (a predecesor of `mlr3`) and `tidymodels`. 
For each model we generate counterfactuals for the first row of the BostonHousing dataset 
using the `WhatIf` method 

```{r}
data(BostonHousing, package = "mlbench")
x_interest = BostonHousing[1L,]
```


## rpart - caret package
```{r}
library("caret")

treecaret = caret::train(medv ~ ., data = BostonHousing[-1,], method = "rpart", 
  tuneGrid = data.frame(cp = 0.01))
predcaret = Predictor$new(model = treecaret, data = BostonHousing[-1L,], y = "medv")
predcaret$predict(x_interest)
nicecaret = NICERegr$new(predcaret, optimization = "plausibility", 
  margin_correct = 0.5, return_multiple = FALSE)
nicecaret$find_counterfactuals(x_interest, desired_outcome = c(30, 40))
```

## rpart - tidymodels package
```{r}
library("tidymodels")
treetm = decision_tree(mode = "regression", engine = "rpart") %>% 
  fit(medv ~ ., data = BostonHousing[-1L,])
predtm = Predictor$new(model = treetm, data = BostonHousing[-1L,], y = "medv")
predtm$predict(x_interest)
nicetm = NICERegr$new(predtm, optimization = "plausibility", 
  margin_correct = 0.5, return_multiple = FALSE)
nicetm$find_counterfactuals(x_interest, desired_outcome = c(30, 40))
```

## rpart - mlr package
```{r}
library("mlr")
task = mlr::makeRegrTask(data = BostonHousing[-1L,], target = "medv")
mod = makeLearner("regr.rpart")

treemlr = train(mod, task)
predmlr = Predictor$new(model = treemlr, data = BostonHousing[-1L,], y = "medv")
predmlr$predict(x_interest)
nicemlr = NICERegr$new(predmlr, optimization = "plausibility", 
  margin_correct = 0.5, return_multiple = FALSE)
nicemlr$find_counterfactuals(x_interest, desired_outcome = c(30, 40))
```

## Decision tree - rpart package 
```{r}
treerpart = rpart(medv ~ ., data = BostonHousing[-1L,])
predrpart = Predictor$new(model = treerpart, data = BostonHousing[-1L,], y = "medv")
predrpart$predict(x_interest)
nicerpart = NICERegr$new(predrpart, optimization = "plausibility", 
  margin_correct = 0.5, return_multiple = FALSE)
nicerpart$find_counterfactuals(x_interest, desired_outcome = c(30, 40))
```

